---
title: "Final Response to the Assignment"
author: "Komla"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
---

# Introduction

This document presents the analysis of the dataset contained in the Excel file **stats_R_Python_questions_2.xlsx**. The assignment consists of three main parts:

1.  **Summary Statistics and Plotting:**\
    Compute the mean and standard deviation of the digital counts and plot these values (Mean ± SD) over time for each marker and treatment group.

2.  **T-Test Comparisons (DAY1 vs. DAY8):**\
    Test whether DAY8 and DAY1 significantly differ at α = 0.05 under each treatment for each marker. A paired t-test is used when the same subjects are measured at both time points; otherwise, an unpaired t-test is applied.

3.  **Mixed-Effects Modeling, Residual Analysis, and Post Hoc Comparisons:**\
    Assess the constant variance assumption via a residuals vs. fitted plot, fit mixed-effects models for each marker, and report both the unadjusted and Bonferroni-adjusted p-values for the DAY22 vs. DAY8 contrast.

Below, each step is described with the corresponding code.

# 1. Summary Statistics and Plotting

## 1.a) Compute Mean and Standard Deviation

```{r summary-stats, echo=TRUE, message=FALSE, warning=FALSE}
# Load necessary libraries
library(readxl)
library(tidyverse)
library(ggplot2)
library(lme4)
library(lmerTest)
library(broom)
library(broom.mixed)
library(emmeans)
library(knitr)  # for kable()

# Set the working directory (adjust this )
setwd("/Users/komlagnona/Desktop/MYDESK/MYWORK")

# Define file path (adjust this)
file_path <- "/Users/komlagnona/Desktop/MYDESK/MYWORK/stats_R_Python_questions_2.xlsx"

# Load the "data" sheet
data <- read_excel(file_path, sheet = "data")

# Check data structure and summary statistics
str(data)
summary(data)

# Compute mean and standard deviation for digital_count by marker, sample_time, and trt_group
summary_stats <- data %>%
  group_by(marker, sample_time, trt_group) %>%
  summarise(
    mean_digital_count = mean(digital_count, na.rm = TRUE),
    sd_digital_count = sd(digital_count, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  as.data.frame()

# View the results in a nicely formatted table
knitr::kable(summary_stats, caption = "Summary Statistics: Mean and SD by Marker, Sample Time, and Treatment Group")
```

# 1.b) Plot Mean ± SD Over Time for Each Marker and Treatment

```{r,fig.height=12, fig.width=15}
# Convert sample_time to an ordered factor before plotting
summary_stats <- summary_stats %>%
  mutate(sample_time = factor(sample_time, levels = c("DAY1", "DAY8", "DAY15", "DAY22", "DAY29")))

# Create the plot using ggplot2
summary_plots <- ggplot(summary_stats, aes(x = sample_time, y = mean_digital_count, 
                                            color = trt_group, group = trt_group)) +
  geom_line(linewidth = 1) +  
  geom_point(size = 2) +  
  geom_errorbar(aes(ymin = mean_digital_count - sd_digital_count, 
                    ymax = mean_digital_count + sd_digital_count), width = 0.2) +  
  facet_wrap(~ marker, scales = "fixed") +
  labs(title = "Marker Readout Over Time",
       x = "Sample Time", 
       y = "Mean Digital Count",
       color = "Treatment Group") +
  scale_color_brewer(palette = "Dark2") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5, color = "black"), # Centered title
    axis.title.x = element_text(size = 18, color = "black"),  # X-axis label
    axis.title.y = element_text(size = 18, color = "black"),  # Y-axis label
    axis.text.x = element_text(size = 14, face = "bold",color = "black"),   # X-axis tick labels
    axis.text.y = element_text(size = 14, face = "bold",color = "black"),   # Y-axis tick labels
    legend.title = element_text(size = 16, color = "black"),  # Legend title
    legend.text = element_text(size = 16, color = "black"),   # Legend text
    strip.text = element_text(size = 14, face = "bold", color = "black")  # Facet titles
  )

# View the plot
print(summary_plots)
```

#2. T-Test Comparisons (DAY1 vs. DAY8)

Let's test whether DAY8 and DAY1 significantly differ under each treatment for each marker. I will use a paired t-test if the same subjects are measured at both time points, and an unpaired t-test otherwise.

```{r}
# Filter data for DAY1 and DAY8 only
day1_day8_data <- data %>% 
  filter(sample_time %in% c("DAY1", "DAY8"))

# Perform t-tests for each marker and treatment group
t_test_results <- day1_day8_data %>%
  group_by(marker, trt_group) %>%
  summarise(
    test = list({
      df <- pick(everything())
      # Identify unique subject IDs for each day
      subs_day1 <- unique(df$sbj_id[df$sample_time == "DAY1"])
      subs_day8 <- unique(df$sbj_id[df$sample_time == "DAY8"])
      
      # Check if the same subjects are present (after sorting)
      if (length(subs_day1) == length(subs_day8) && all(sort(subs_day1) == sort(subs_day8))) {
        # For paired t-test: sort by subject ID to ensure proper pairing
        x_vals <- df %>% 
          filter(sample_time == "DAY1") %>% 
          arrange(sbj_id) %>% 
          pull(digital_count)
        y_vals <- df %>% 
          filter(sample_time == "DAY8") %>% 
          arrange(sbj_id) %>% 
          pull(digital_count)
        
        t.test(x = x_vals, y = y_vals, paired = TRUE)
      } else {
        # Otherwise, use an unpaired t-test
        x_vals <- df %>% filter(sample_time == "DAY1") %>% pull(digital_count)
        y_vals <- df %>% filter(sample_time == "DAY8") %>% pull(digital_count)
        
        t.test(x = x_vals, y = y_vals, paired = FALSE)
      }
    }),
    .groups = "drop"
  ) %>%
  mutate(
    results = map(test, tidy)
  ) %>%
  unnest(results) %>%
  select(marker, trt_group, p.value, estimate, statistic)

# View the t-test results in a nicely formatted table
knitr::kable(t_test_results, caption = "T-Test Results (DAY1 vs. DAY8) by Marker and Treatment Group")

# Select only significant results (p < 0.05)
significant_t_test_results <- filter(t_test_results, p.value < 0.05)
selected_markers           <- significant_t_test_results$marker
selected_trt_group         <- significant_t_test_results$trt_group

# Print summary statement for significant markers
cat("DAY1 and DAY8 differ for the following marker(s):", 
    selected_markers, "in the treatment group(s):", selected_trt_group, "\n")


```

***Summary.***

-   The difference in treatments between DAY8 and DAY1 is significantly different (P \< 0.05) for marker C4 under treatment TA. Note: the evidence is suggestive (p \< 0.1) for marker TG under treatments TA and TB

#3. Mixed-Effects Modeling, Residual Analysis, and Post Hoc Comparisons

## 3.a) Residuals vs. Fitted Plot

First, let's assess any departure from the assumptions unlying the tukey test. We already knowthe experimental units are randomly drawn from a population. That said, let's check for the constant variance (homoscedasticity) assumption by plotting the residuals versus the fitted values for each marker.

```{r,fig.height=12, fig.width=15}
# Function to fit the model and append residuals and fitted values for a given marker
analyze_marker_resid <- function(df) {
  model <- lmer(digital_count ~ sample_time + trt_group + (1 | sbj_id), data = df)
  df %>% mutate(fitted = fitted(model), residuals = resid(model))
}

# Apply the function to each marker and combine the results
resid_data <- data %>%
  group_by(marker) %>%
  group_modify(~ analyze_marker_resid(.x)) %>%
  ungroup()

# Create the Residuals vs. Fitted plot for all markers
p <- ggplot(resid_data, aes(x = fitted, y = residuals)) +
  geom_point(color = "blue") +
  geom_hline(yintercept = 0, color = "black", linetype = "dashed") +
  facet_wrap(~ marker, scales = "free") +
  labs(title = "Residuals vs Fitted Values by Marker",
       x = "Fitted Values",
       y = "Residuals") +
  theme_minimal()

#
p<- p + theme(text = element_text(size = 20), plot.title = element_text(size = 16, face = "bold", hjust = 0.5, color = "black"))
# Visualize the plot
print(p)


```

***Summary.***

-   The constant variance assumption appears to be reasonable for some markers than the others.For instance, I suspect heteroskedasticity for marker C4. That said, I will not use the Turkey test for inference here. I will report the unadjusted p-values along with the corresponding adjusted p-values after multiple testing correction.

## 3.b) Mixed-Effects Model with Post Hoc Comparisons

Here, i fit a mixed-effects model for each marker, extract the treatment effect p-values, and perform pairwise comparisons for the contrast between DAY22 and DAY8. Both unadjusted and Bonferroni-adjusted p-values are reported.

```{r}
marker_results <- data %>%
  group_by(marker) %>%
  nest() %>%
  mutate(
    # Fit the model for each marker
    model = map(data, ~ lmer(digital_count ~ sample_time + trt_group + (1 | sbj_id), data = .x)),
    
    # Extract the ANOVA table and get the p-value for the treatment effect
    anova_out = map(model, ~ anova(.x)),
    trt_p_value = map_dbl(anova_out, ~ .x["trt_group", "Pr(>F)"]),
    
    # Compute estimated marginal means for sample_time
    emms = map(model, ~ emmeans(.x, ~ sample_time)),
    
    # Get unadjusted pairwise comparisons among sample_time levels
    pairwise_unadj = map(emms, ~ contrast(.x, method = "pairwise", adjust = "none")),
    
    # Get Bonferroni-adjusted pairwise comparisons among sample_time levels
    pairwise_bonf = map(emms, ~ contrast(.x, method = "pairwise", adjust = "bonferroni")),
    
    # Extract the unadjusted p-value for the contrast between DAY22 and DAY8
    day22_day8_p_unadj = map_dbl(pairwise_unadj, ~ {
      tmp <- as.data.frame(.x) %>% filter(contrast %in% c("DAY22 - DAY8", "DAY8 - DAY22"))
      if(nrow(tmp) > 0) tmp$p.value[1] else NA_real_
    }),
    
    # Extract the Bonferroni-adjusted p-value for the contrast between DAY22 and DAY8
    day22_day8_p_bonf = map_dbl(pairwise_bonf, ~ {
      tmp <- as.data.frame(.x) %>% filter(contrast %in% c("DAY22 - DAY8", "DAY8 - DAY22")) #making sure the proper test is extracted. DAY 22-DAY8 is equivalent of DAY8 - DAY22                                         
      if(nrow(tmp) > 0) tmp$p.value[1] else NA_real_
    })
  ) %>%
  select(marker, trt_p_value, day22_day8_p_unadj, day22_day8_p_bonf)

# View the mixed model results in a nicely formatted table
knitr::kable(marker_results, caption = "Mixed-Effects Model Results with Unadjusted and Bonferroni-adjusted p-values")


```

***Summary.***

Here, I report the unadjusted and adjusted p-values. I opted to use Bonferroni correction to adjust for multiple testing and to control for the family wise error rate.

-   The p-values for the treatment effect of all markers are \> 0.05. In other words, there is not enough statistical evidence to conclude that the treatment (trt_group) has an effect on the marker digital_count at the 5% significance level

-   Aslo, we do not have enough evidence to conclude that the difference in digital_count between DAY22 and DAY8 for all markers is significant (p \> 0.05).

-   Note: the evidence is suggestive (p-val \< 0.1) for markers C4 & C8
